---
title: "Karl Task 4"
author: "Taryn H"
date: "2023-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load(tidyverse, tidymodels, lme4, sjPlot)
#install.packages("lme4")
#install.packages("sjPlot")
library("lme4")
library("sjPlot")

```


#Load in data
```{r}
data<-read_csv(here::here("raw-data", "Gene_Expression.csv"))
#data <- data |> select(Cell_Line, Concentration, Treatment,Gene_Expression)

data <- data |> filter(Gene_Expression>0)
data
```

#Random Intercept Model - Treatment as Random Effect
```{r}
model1<-lmer(Gene_Expression ~ Concentration + 
                       Cell_Line +
                       (1 | Treatment), data=data)
summary(model1)

coef(summary(model1))
```

#Random Intercept Model - Random Slope for Treatment
```{r}
model2<-lmer(Gene_Expression ~  Cell_Line + (0 + Concentration | Treatment), data=data)


summary(model2)

coef(summary(model2))

```

#Random Slope and Intercepts for Treatment
```{r}
model5<-lmer(Gene_Expression ~ Concentration + 
                       (1 + Concentration | Treatment), data=data)


summary(model5)

coef(summary(model5))
```


#Random Intercept Model - Cell Line as Random Effect
```{r}
model3<-lmer(Gene_Expression~Concentration + 
                        Treatment +
                        (1 | Cell_Line), data=data)

summary(model3)

coef(summary(model3))
```

#Random Intercept Model - Random Slope for Cell Line
```{r}
model4<-lmer(Gene_Expression ~  Treatment + (0 + Concentration | Cell_Line), data=data)

summary(model4)

coef(summary(model4))
```



#AIC Comparison
```{r}
AIC(model1, model2, model3, model4)
```

#Producing visualisation for model 3
```{r}
ranef(model3)

fixef(model3)

#sjp.lmer(model3, vars="Concentration", type="ri.slope")

data |> mutate(my_model = predict(model3)) |>
  ggplot(aes(x=Concentration, y=Gene_Expression, col=`Cell Line_Treatment`)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model))


```

#Producing visualisation for model 2
```{r}

data |> mutate(my_model = predict(model2)) |>
  ggplot(aes(x=Concentration, y=Gene_Expression, col=`Cell Line_Treatment`)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model))

```



#Stuff I may not need
```{r}
data1 <- data |> filter(Cell_Line == "Wild Type" & Treatment == "Placebo")
plot1<- data1 |> mutate(my_model_new=predict(model3, newdata=data1)) |> 
  ggplot(aes(x=Concentration, y=Gene_Expression, col=Cell_Line)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model_new)) + 
  scale_fill_discrete(labels=c("Wild Type and Placebo"))
  

data2 <- data |> filter(Cell_Line == "Wild Type" & Treatment == "Activating Factor 42")
plot2<-data2 |> mutate(my_model_new=predict(model3, newdata=data2)) |> 
  ggplot(aes(x=Concentration, y=Gene_Expression, col=Treatment)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model_new))

data3 <- data |> filter(Cell_Line == "Cell Type 101" & Treatment == "Placebo")
plot3<-data3 |> mutate(my_model_new=predict(model3, newdata=data3)) |> 
  ggplot(aes(x=Concentration, y=Gene_Expression, col=Cell_Line)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model_new))

data4 <- data |> filter(Cell_Line == "Cell Type 101" & Treatment == "Activating Factor 42")
plot4<-data4 |> mutate(my_model_new=predict(model3, newdata=data4)) |> 
  ggplot(aes(x=Concentration, y=Gene_Expression, col=Cell_Line)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model_new))

plot1 + plot2 + plot3 + plot4

data1 <- data |> filter(Cell_Line == "Wild Type" & Treatment == "Placebo")
plot1<- data1 |> mutate(my_model_new=predict(model3, newdata=data1)) |> 
  ggplot(aes(x=Concentration, y=Gene_Expression, col=Cell_Line)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model_new)) + 
  scale_fill_discrete(labels=c("Wild Type and Placebo"))
  

data2 <- data |> filter(Cell_Line == "Wild Type" & Treatment == "Activating Factor 42")
plot2<-data2 |> mutate(my_model_new=predict(model3, newdata=data2)) |> 
  ggplot(aes(x=Concentration, y=Gene_Expression, col=Treatment)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model_new))

data3 <- data |> filter(Cell_Line == "Cell Type 101" & Treatment == "Placebo")
plot3<-data3 |> mutate(my_model_new=predict(model3, newdata=data3)) |> 
  ggplot(aes(x=Concentration, y=Gene_Expression, col=Cell_Line)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model_new))

data4 <- data |> filter(Cell_Line == "Cell Type 101" & Treatment == "Activating Factor 42")
plot4<-data4 |> mutate(my_model_new=predict(model3, newdata=data4)) |> 
  ggplot(aes(x=Concentration, y=Gene_Expression, col=Cell_Line)) + geom_point() +
  geom_line(aes(x=Concentration, y=my_model_new))

plot1 + plot2 + plot3 + plot4

```

